---
title: "Thurstonian IRT Example"
author: ""
output:
  html_document:
    css: styles.css
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    code_folding: show
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Thurstonian IRT Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE,warning=FALSE,message=FALSE}
directory<-if (rstudioapi::isAvailable()) paste0(dirname(rstudioapi::getActiveDocumentContext()$path),"/") else ""
source(paste0(directory, "R/code_options.R"))
source(paste0(directory, "R/code_rwf.R"))

```

## Load

```{r library,message=FALSE,warning=FALSE}
library(thurstonianIRT)
```

The model is the Thurstonian IRT model of Brown and Maydeu-Olivares (2011), fitted with the thurstonianIRT package (Bürkner, Schulte & Holling, 2019).

## The data

Forced-choice rankings must be expressed as binary pairwise outcomes: for each pair {i, k} in a block, y = 1 if item i is preferred to item k and 0 otherwise (Brown & Maydeu-Olivares, 2011).
In the examples below, rank 1 denotes the most preferred item.
The `triplets` data are already in this pairwise format; raw ranks could instead be passed to `make_TIRT_data(..., format = "ranks", direction = "smaller")`, which performs this recoding automatically.

### Example 1

| Rank | Rank | Rank | 
|------|------|------|
| A    | B    | C    |
| 1    | 2    | 3    |

| binary | binary | binary | 
|--------|--------|--------|
| {A, B} | {A, C} | {B, C} |
|   1    |    1   |   1    |


### Example 2

| Rank | Rank | Rank | 
|------|------|------|
| A    | B    | C    |
| 2    | 1    | 3    |

| binary | binary | binary | 
|--------|--------|--------|
| {A, B} | {A, C} | {B, C} |
|   0    |    1   |   1    |

## The coded data

```{r load}
data("triplets")
head(triplets)
```

## Define blocks

```{r blocks}
blocks<-set_block(c("i1","i2","i3"),
                  traits=c("t1","t2","t3"),
                  signs=c(1,1,1))+  
  set_block(c("i4","i5","i6"),
            traits=c("t1","t2","t3"),
            signs=c(-1,1,1))+
  set_block(c("i7","i8","i9"),
            traits=c("t1","t2","t3"),
            signs=c(1,1,-1))+
  set_block(c("i10","i11","i12"),
            traits=c("t1","t2","t3"),
            signs=c(1,-1,1))
data.frame(blocks$blocks)
```

## Triplets

```{r triplets}
data_tirt<-make_TIRT_data(data=triplets,blocks=blocks,direction="larger",format="pairwise",family="bernoulli",range=c(0,1))

data_sem<-make_sem_data(data=data_tirt)
data_stan<-make_stan_data(data=data_tirt)

spesification_lavaan<-make_lavaan_code(data_tirt)
spesification_mplus<-make_mplus_code(data_tirt)

```

## Model Specification
```{r spesification}
# Specification in lavaan
cat(spesification_lavaan)

# Specification in MPLUS
cat(spesification_mplus$TITLE)
cat(spesification_mplus$DATA)
cat(spesification_mplus$VARIABLE)
cat(spesification_mplus$ANALYSIS)
cat(spesification_mplus$MODEL)
cat(spesification_mplus$MODELCONSTRAINT)
cat(spesification_mplus$SAVEDATA)

```


## Model fit

```{r fit, cache=TRUE}
fit_lavaan<-fit_TIRT_lavaan(data_tirt,control=list(maxiter=1000))
# fit_mplus<-fit_TIRT_mplus(data_tirt)
fit_stan<-fit_TIRT_stan(data_tirt)
model<-fit_lavaan$fit 

predict(fit_lavaan,newdata=data_tirt)

```

## Plot Model

```{r plot,fig.width=10,fig.height=10}
semPlot::semPaths(model,what="est",rotation=2,reorder=FALSE)
semPlot::semPaths(model,what="std",rotation=2,reorder=FALSE)
semPlot::semPaths(model,what="paths",rotation=2,reorder=FALSE)
```

## Evaluate Model
```{r evaluate}
# Fit indices (ULSMV estimator: robust chi-square-based indices; AIC/BIC are not available)
fit_index<-data.frame(fit=lavaan::fitMeasures(model,c("chisq.scaled","df.scaled","pvalue.scaled","cfi.scaled","tli.scaled","rmsea.scaled","srmr","gfi")))
round(fit_index,2)

# Explained variance
r_squared<-data.frame(r_squared=lavaan::inspect(model,"rsquare"))
r_squared

# Estimates
unstandardized_estimates<-lavaan::inspect(model,"est")
unstandardized_estimates
standardized_estimates<-lavaan::inspect(model,"std")
standardized_estimates

lavaan::parameterEstimates(model,se=TRUE,
                           zstat=TRUE,
                           pvalue=TRUE,
                           ci=TRUE,
                           level=0.95,
                           standardized=TRUE,
                           fmi=FALSE,
                           remove.system.eq=TRUE,
                           remove.eq=FALSE,
                           remove.ineq=FALSE,
                           remove.def=FALSE,
                           rsquare=TRUE,
                           add.attributes=TRUE)
# Modification indices
lavaan::modificationIndices(model,
                            standardized=TRUE,
                            cov.std=TRUE,
                            information="expected",
                            power=TRUE,
                            delta=0.1,
                            alpha=0.05,
                            high.power=0.75,
                            sort.=TRUE,
                            minimum.value=0,
                            free.remove=FALSE,
                            na.remove=TRUE,
                            op=NULL)

```


## Score data

```{r score,message=FALSE}
score<-predict(fit_lavaan)
score<-data.frame(score)
score<-reshape(score, direction="wide", idvar="id", timevar="trait")

triplets_new<-make_TIRT_data(data=triplets[1:2,],blocks=blocks,direction="larger",format="pairwise",family="bernoulli",range=c(0,1))
score_new<-predict(fit_stan,newdata=triplets_new)
score_new<-data.frame(score_new)
score_new<-reshape(score_new[,c("id","trait","estimate")], direction="wide", idvar="id", timevar="trait")

```

## Print Scores

`score` contains lavaan (ULSMV) factor-score estimates for all `r nrow(score)` respondents; `score_new` contains Bayesian (Stan) posterior-mean estimates for two respondents treated as new data, so small differences between the two are expected.

```{r score_new}

head(score)

score_new

```

## References

Brown, A., & Maydeu-Olivares, A. (2011). Item response modeling of forced-choice questionnaires. *Educational and Psychological Measurement, 71*(3), 460–502. https://doi.org/10.1177/0013164410375112

Bürkner, P.-C., Schulte, N., & Holling, H. (2019). On the statistical and practical limitations of Thurstonian IRT models. *Educational and Psychological Measurement, 79*(5), 827–854. https://doi.org/10.1177/0013164419832063










